\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-05-06 honors_thesis/journal_paper/paper.tex
Finished a draft of the final version of the journal paper.  One more pass...

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
rest of this paper that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.

@@

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
has a global optimum with a value of 0, and sometimes the swarm finds the
optimum and sometimes it doesn't.  Instead of showing average function value at
each iteration, a more informative plot for Griewank shows the percent of runs
that have found the global optimum by each iteration.  We show results in
\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO algorithm finds the optimum faster
than the original PSO.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank2}
\end{figure}

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum twice as fast
as the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank3}
\end{figure}

Sphere is a function for which very little exploration needs to be done, so
always keeping the best speculative position found makes intuitive sense.  A
less intuitive result occurs when using Pick Best with the function Griewank.
Griewank is a highly deceptive function prone to premature convergence.  Our
intuition on the behavior of picking the best speculative child is that it
would tend to prematurely converge more often than the original PSO.  However,
as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
the optimum while also reducing the number of iterations required to converge.
Perhaps Pick Best encourages local exploitation, while the sparse communication
of the Ring topology still allows for global exploration.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

When looking at the function Griewank, presented in \fig{griewank-pruned}, we
see different results.  Our intuition about picking the best child is more
correct in this case; while adding 50 more particles to the swarm, Pick Best
with pruning still gets stuck more often than the original Pick Best.  However,
Social Promotion does well in this case, finding the optimum every time and
faster than the original PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

With the function Griewank, the premature convergence problems inherent in
picking the best child are exacerbated when speculating several iterations
ahead.  When the algorithm finds the optimum, it finds it quicker than any
other method we tried.  However, it also gets stuck and fails to find the
optimum more than any other method.  In order for this method to be
consistently useful on functions like Griewank, it seems that more processors
are required.  The results are shown in \fig{griewank-manyiters}.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

\subsection{Bohachevksy}

@@

@@

Griewank is a function in between Sphere and Rastrigin.  It is highly deceptive
and is prone to premature convergence, but by adding particles to the swarm a
point is reached where ``enough'' exploration is done, and the algorithm finds
the optimum essentially all of the time.  For such functions, the best approach
seems to be to increase the swarm size until ``enough'' exploration is reached,
then use extra processors to perform speculative evaluation and increase the
number of iterations performed.  Sphere and Rastrigin can be thought of as
special cases of these types of functions; Sphere simply needs a very small
swarm size to produce ``enough'' exploration, and Rastrigin requires a very
large swarm.  We expect that for all functions there is a swarm size for which
additional particles are less useful than additional iterations.

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.

@@

\bibliographystyle{unsrt}
\bibliography{%
../../../bib/aml/bib,%
../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/sim-anneal/parallel/bib,%
../../../bib/ga/bib}

>>>>

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
\sec{results} that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited 
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to perform as well as standard PSO because of their smaller swarm size.
In comparing SEPSO to standard PSO in \fig{schwefel1}, we can see that SEPSO
fails to perform any better than PSO with a Random topology, and it is by far
outperformed by PSO with a Complete topology.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel1}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:schwefel1}
\end{figure}

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than PSO Complete, with Many Iterations performing comparably to
PSO.  The only algorithm that performed better than PSO was Pick Best Pruned,
likely because of the larger swarm size available to it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel2}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations, comparing all speculative algorithms.}
  \label{fig:schwefel2}
\end{figure}

@@

Next we look at the function Griewank.  It is generally recommended to use the
Ring topology when optimizing the Griewank function, as Complete is prone to
premature convergence on a local
optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank has a global
optimum with a value of 0, and sometimes the swarm finds the optimum and
sometimes it doesn't.  Instead of showing average function value at each
iteration, a more informative plot for Griewank shows the percent of runs that
have found the global optimum by each iteration.  We ran 50 trials of each
experiment with Griewank, so that the curves are more smooth.  We show results
in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO algorithm finds the optimum faster
than the original PSO.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank1}
\end{figure}

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank2}
\end{figure}

In order to show the premature convergence properties of the various algorithms
we have presented, we will show results first with swarms using 240 processors,
where these are more visible.  Then we will present an overall result for 800
processors, where using our techniques is generally more successful.

When we look at the performance of our Pick Best algorithm, a less intuitive
result occurs with Griewank than with Sphere.  Because Sphere needs very little
exploration, the seemingly greedy approach of Pick Best does very well.  But it
also greatly improves performance on Griewank.  In \fig{griewank-pickbest} we
see that Pick Best improves accuracy over SEPSO by 20\%, while at the same time
finding the optimum over 100 rounds of evaluation sooner on average.  It seems
that while Pick Best is locally greedy, there is enough exploration in the
seven speculative evaluations to overcome the inherent greediness of the
algorithm.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

When we introduce pruning, our intuition about Pick Best turns out to be more
correct.  While adding 50 more particles to the swarm, Pick Best with pruning
still gets stuck just as often as the original Pick Best.  However, Social
Promotion does well with pruning; it increases the success rate to close to
100\%, while still finding the optimum on average much faster than the original
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

With Griewank, the premature convergence problems inherent in picking the best
child are exacerbated when speculating several iterations ahead.  When Many
Iterations finds the optimum, it finds it quicker than any other method we
tried, on average four times faster than standard PSO.  However, it also gets
stuck and fails to find the optimum more than any other method.  The results
are shown in \fig{griewank-manyiters}. (TODO: Rerun Many Iterations with the
right swarm size)

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

Now that we have seen the effect that the various algorithms have on speed of
convergence and success rates, we show a result for using 800 processors, where
most of the algorithms have a 100\% success rate.  In this case, the best
algorithm to use is the one which finds the optimum the fastest.  As can be
seen in \fig{griewank-800}, with 800 processors the best choice is Pick Best.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank-800}
\end{figure}

\subsection{Bohachevsky}

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes doesn't.  Since we have already
demonstrated the convergence properties of the various algorithms with
Griewank, for Bohachevsky we merely present the best results.  In
\fig{bohachevsky} we can see that all of the speculative algorithms found the
optimum much quicker than PSO with a Random topology.  However, SEPSO was
slower than PSO Complete and got stuck around 15\% of the time.  Using Pick
Best, Pick Best Pruned, and Many Iterations all outperformed PSO Complete,
though Many Iterations only almost reached a 100\% success rate.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with a swarm that uses 800 processors per round
  of evaluations.}
  \label{fig:bohachevsky}
\end{figure}

@@

We have given five different variants of speculative evaluation, each of which
has different properties.  The variants of the algorithm perform differently on
different functions and at different swarm sizes, as would be expected by the
No Free Lunch Theorem~\cite{wolpert-1997-nfl-for-optimization}.  We have given
a brief evaluation of the premature convergence properties of these variants on
deceptive functions.  Though there is no one variant that performs best across
all functions, for all but one of the functions we looked at \emph{some} kind
of speculative evaluation outperforms standard PSO, often dramatically.

@@

Griewank is a function in between Sphere and Rastrigin (as are Bohachevsky and
Schwefel 2.21).  It is highly deceptive and is prone to premature convergence,
but by adding particles to the swarm a point is reached where ``enough''
exploration is done, and the algorithm finds the optimum essentially all of the
time.  For such functions, the best approach seems to be to increase the swarm
size until ``enough'' exploration is reached, then use extra processors to
perform speculative evaluation and increase the number of iterations performed.
Sphere and Rastrigin can be thought of as special cases of these types of
functions; Sphere simply needs a very small swarm size to produce ``enough''
exploration, and Rastrigin requires a very large swarm.  We expect that for all
functions there is a swarm size for which additional particles are less useful
than additional iterations.

@@

@@

\bibliographystyle{unsrt}
\bibliography{%
../../../bib/aml/bib,%
../../../bib/nfl/bib,%
../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/sim-anneal/parallel/bib,%
../../../bib/ga/bib}
